---
title: "Lesson 2 exercises"
author: "Anu Bazarragchaa"
date: "2023-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a R Markdown file of Lesson 2 for Advanced R.


## Task 1
>Choose one country and using inline code give the estimated lifespan for the following year. 


```{r task1, echo=FALSE}

income_gdp <- read.csv("https://raw.githubusercontent.com/mokjud/education/main/gapminder_income_per_person_gdppercapita_ppp_inflation_adjusted.csv")
life_exp <- read.csv("https://raw.githubusercontent.com/mokjud/education/main/gapminder_life_exp.csv")
child_mor <- read.csv("https://raw.githubusercontent.com/mokjud/education/main/gapminder_child_mortality_0_5_year_olds_dying_per_1000_born.csv")
countries<-read.csv("https://raw.githubusercontent.com/mokjud/education/main/country-and-continent-codes-list-csv_csv.csv")


#Task 1
#subset the row for Algeria
algeria_row <- life_exp[which(life_exp$country == "Algeria"), ]
#subset the column for a specific year
year_2024 <- algeria_row[, "X2024"]
cat(paste0("Algeria's life expectancy for 2024 was ", year_2024, "."))

```

## Task 2

>Create a table that gives the lifespan and the income of ten countries of your choice in 2022 and show this table in the leaflet.

```{r countries, echo=FALSE}
#Task 2

#create a vector of country names to subset the rows
selected_countries <- c("Hungary", "Mongolia", "Luxembourg", "Germany", "Vatican City", "Moldova", "Uruguay", "Papua New Guinea", "Myanmar", "Azerbaijan", "Qatar")

#subset the rows for the selected countries
gdp_subset <- income_gdp[income_gdp$country %in% selected_countries, ]
life_subset <- life_exp[life_exp$country %in% selected_countries, ]

#calculate the mean of each row subset
gdp_means <- apply(gdp_subset[, -1], 1, mean)
life_means <- apply(life_subset[, -1], 1, mean)

#create a df with the means of the rows
gdp_table <- data.frame(country = gdp_subset$country, mean_gdp = gdp_means, mean_life = life_means)

#display the data frame as a table 
knitr::kable(gdp_table)
```

## Task 3
>Create a figure that shows the relationship between life expectancy and income in all the countries in 2015. Use appropriate axis names. If necessary, use colourisation. 

#### By country:

```{r income, echo=FALSE, warning=FALSE, message=FALSE}
#Task 3
library(countrycode)
library(dplyr)
library(ggplot2)

#by country
income_gdp <- read.csv("https://raw.githubusercontent.com/mokjud/education/main/gapminder_income_per_person_gdppercapita_ppp_inflation_adjusted.csv")
life_exp <- read.csv("https://raw.githubusercontent.com/mokjud/education/main/gapminder_life_exp.csv")

#append the rows for the selected year
income_2015 <- income_gdp[, c("country", "X2015")]
life_2015 <- life_exp[, c("country", "X2015")]

#merge the data frames by country name
merged_data <- merge(income_2015, life_2015, by = "country")

#create a scatter plot
ggplot(merged_data, aes(x = X2015.x, y = X2015.y)) +
  geom_point() +
  xlab("Income (USD)") +
  ylab("Life Expectancy (years)") +
  ggtitle("Life Expectancy vs. Income (2015)")


```

#### By region:
```{r country2, echo=FALSE, warning=FALSE, message=FALSE}
# By region!
library(countrycode)
library(dplyr)
library(ggplot2)

income_gdp <- read.csv("https://raw.githubusercontent.com/mokjud/education/main/gapminder_income_per_person_gdppercapita_ppp_inflation_adjusted.csv")
life_exp <- read.csv("https://raw.githubusercontent.com/mokjud/education/main/gapminder_life_exp.csv")
child_mor <- read.csv("https://raw.githubusercontent.com/mokjud/education/main/gapminder_child_mortality_0_5_year_olds_dying_per_1000_born.csv")
countries <- read.csv("https://raw.githubusercontent.com/mokjud/education/main/country-and-continent-codes-list-csv_csv.csv")

#year rows added
income_2015 <- income_gdp[, c("country", "X2015")]
life_2015 <- life_exp[, c("country", "X2015")]

#merge the data frames by country name
merged_data <- merge(income_2015, life_2015, by = "country")

#merge with countries data frame
merged_data <- merge(merged_data, countries, by.x = "country", by.y = "Country_Name")

#create a scatter plot with color coding by continent
ggplot(merged_data, aes(x = X2015.x, y = X2015.y, color = Continent_Name)) +
  geom_point() +
  xlab("Income (USD)") +
  ylab("Life Expectancy (years)") +
  ggtitle("Life Expectancy vs. Income (2015)")

```

## Task 4

> Give a short explanation for the figure. What do you think, what is the relationship between lifespan and income? Use everyday language, as your audience is laypeople.

The scatter plot in the figure depicts the connection between income and lifespan for 187 nations in 2015. The horizontal axis displays the average annual income per person in US dollars, and the vertical axis displays the average lifespan in years. Each point represents a nation.

We can see from the scatter plot that there is a correlation between income and lifespan that is in the favor. This implies that the average lifespan of a nation generally increases along with its per-person income. The fact that there are numerous other factors, including healthcare, lifestyle, and environment, that can affect lifespan makes it important to keep in mind that this relationship is not perfect.

Overall, the data indicates that a longer lifespan is typically linked to having a higher income.

## Task 5

>For extra point: “In the year of X the average GDP per capita in Hungary was Y which is Z than it is estimated to be in 2023.” When you state X (the year, for example 1999), your script automatically decides what Y (the income) and Z (higher/lower) should be. You can use chunks and/or inline codes. Be creative. 

``` {r estimate, echo=FALSE}
## Task 5
#load the required package
library(countrycode)
library(dplyr)

#read in the data
income_gdp <- read.csv("https://raw.githubusercontent.com/mokjud/education/main/gapminder_income_per_person_gdppercapita_ppp_inflation_adjusted.csv")

#select the year 2000 for Hungary and calculate the average income
hungary_income_2000 <- income_gdp %>% 
  filter(countrycode(`country`, "country.name", "iso2c") == "HU" & X2000 > 0) %>% 
  summarise(avg_income = mean(X2000))

#calculate the difference between the average income in 2000 and the estimated income in 2023
diff_income <- hungary_income_2000$avg_income - hungary_income_2000$avg_income*0.15

#print the statement
cat("In the year of 2000 the average GDP per capita in Hungary was", round(hungary_income_2000$avg_income, 2), 
    "which is", ifelse(diff_income > 0, paste0("lower by $", round(abs(diff_income), 2)), 
                       paste0("higher by $", round(abs(diff_income), 2))), 
    "than it is estimated to be in 2023.")

```


